---
title: "Take-home Exercise 2"
description: |
  Creating data visualisation beyond default: peer critique and visualization remake of take-home exercise 1.
author:
  - name: Leslie Long Nu
    url: https://www.linkedin.com/in/leslielongnu/
    affiliation: School of Computing and Information Systems (SMU)
    affiliation_url: https://scis.smu.edu.sg/
date: "`r Sys.Date()`"
preview:
output:
  distill::distill_article:
    code_folding: hide
    self_contained: false
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r remake site, inlcude= FALSE, eval=FALSE}
https://visualanalytics.netlify.app/th_ex/th_ex1.html
```

## 1. Overview 
This take-home exercise is done based on a take-home exercise 1 submission prepared by a classmate. The peer submission will be critiqued in terms of clarity and aesthetics, and the original design will be remade using the data visualization principles and best practice learnt in Lesson 1 and 2. 

The dataset used in take-home exercise 1 and 2 is downloaded from [VAST Challenge 2022](https://vast-challenge.github.io/2022/), and processed by RStudio `tidyverse` family of packages and visualized by `ggplot2` and its extensions.

## 2. Critique

```{r install package, include=FALSE}
packages = c('tidyverse', 'knitr', 'ggdist', 'scales', 'grid', 'gridExtra',
             'formattable', 'patchwork')
for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```

```{r import data, include=FALSE}
# <details><summary>Code Chunk</summary>
participants <- read_csv('data/Participants.csv')
# </details>
```


```{r, bin age, include=FALSE}
one <- c("18", "19", "20")
two <- c("21", "22", "23", "24", "25", "26", "27", "28", "29", "30" )
three <- c("31", "32", "33", "34", "35", "36", "37", "38", "39", "40")
four <- c( "41", "42", "43", "44", "45", "46", "47", "48", "49", "50")
five <-  c( "51", "52", "53", "54", "55", "56", "57", "58", "59", "60")

participants <- participants %>%
  mutate(age_modified = case_when(
    age %in% one ~ "<=20",
    age %in% two ~ "20's",
    age %in% three ~ "30's",
    age %in% four ~ "40's",
    age %in% five ~ "50+")) %>%
  select(-age)

```

### 2.1 Distribution of Joviality

```{r joviality distribution}
ggplot(data=participants,
       aes(x = joviality)) +
  geom_histogram(bins=20,
                 boundary = 50,
                 color="black",
                 fill="light blue") +
  coord_cartesian(xlim=c(0.01,1)) +
  labs(title = "Distribution of Joviality",
       caption = "demographic information, Ohio USA") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 1),
        plot.caption = element_text(hjust = 0))
```
A histogram is used to reveal the distribution of residents' joviality. 

**Clarity**

**Aesthetics**

## 2.2 Distribution of Age for Different Household Types

```{r age & edu }
ggplot(data=participants, 
       aes(x=age_modified, fill = educationLevel)) +
  geom_bar()+
  labs(
    title = "Distribution of Age for different household types",
    caption = "demographic information, Ohio USA"
    ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 1),
    plot.caption = element_text(hjust = 0)
    )
```
**Clarity**

**Aesthetics**

## 2.2 Joviality Measure

```{r}
ggplot(data=participants, 
       aes(x= joviality, 
           y= educationLevel, fill = haveKids)) +
  geom_col() +
  theme_classic()+
  labs(
    title = "Joviality Measure",
    caption = "demographic information, Ohio USA"
    ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 1),
    plot.caption = element_text(hjust = 0)
    )+
  facet_wrap(~age_modified)
```


**Clarity**

**Aesthetics**

## 2.4

```{r}
ggplot(data=participants, 
       aes(x= interestGroup, 
           y= joviality)) +
  geom_col() +
  theme_classic()+
  labs(
    title = "Joviality Measure",
    caption = "demographic information, Ohio USA"
    ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 1),
    plot.caption = element_text(hjust = 0)
    )+
  facet_wrap(~age_modified)
```


**Clarity**

**Aesthetics**

## 3. Visualization Remake 

## 4. Learning Points 


