---
title: "Take-home Exercise 4"
categories: Take-home Exercise
description: |
  Putting Visual Analytics into Practical Use: 
  VAST Challenge 2022, Challenge 2: Patterns of Life. 
author:
  - name: Leslie Long Nu
    url: https://www.linkedin.com/in/leslielongnu/
    affiliation: SMU, MITB
    affiliation_url: https://scis.smu.edu.sg/master-it-business
date: 2022-05-22
preview:
output:
  distill::distill_article:
    code_folding: show
    self_contained: false
    toc: true
    toc_depth: 3
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

# 1. Overview
## 1.1 Introduction
With reference to Challenge 2 Question 3 of [VAST Challenge 2022](https://vast-challenge.github.io/2022/), this take-home exercise will reveal the daily routines of **two** selected participants of the city of Engagement, Ohio USA. [`ViSiElse`](https://rdrr.io/cran/ViSiElse/man/) and other appropriate visual analytics R packages will be used. 

*Challenge 2: Patterns of Life, Question 3*

*Participants have given permission to have their daily routines captured. Choose two different participants with different routines and describe their daily patterns, with supporting evidence.*

## 1.2 Methodology
**Scope**
As the VAST challenge data include participants' daily routine logs spanning for 15 months, the scope of this take-home exercise is scaled down time limit (1 week) of the take-home exercise, the scope of this take-home exercise is defined as: 

# 2. Data Preparation 
## 2.1 Install and Load Packages 
The following code chunk installs the required R packages and loads them onto RStudio environment. 

```{r load libraries}
packages = c('scales', 'viridis', 'lubridate', 'ggthemes', 
             'gridExtra', 'tidyverse', 'readxl', 'knitr',
             'data.table', 'ViSiElse')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p,character.only = T)
}
```

## 2.2 Import Raw Datasets
Relevant datasets are imported using `read_csv()` of readr package, which is useful for reading delimited files into tibbles. 



## 2.3 Data Wrangling
## 2.4 Save as and Read RDS Files

# 3. Visualizations and Insights

# 4. Learning Points


