---
title: "In-class Exercise 6: Visualizing Geospacial Data"
categories: In-class Exercise 
description: |
  Lesson 6: GeoVisual Analytics, in-class exercise. 
author:
  - name: Leslie Long Nu
    url: https://www.linkedin.com/in/leslielongnu/
    affiliation: SMU, MITB
    affiliation_url: https://scis.smu.edu.sg/master-it-business
date: 2022-05-21
preview:
output:
  distill::distill_article:
    self_contained: false
    code_folding: show
    toc: true
    toc_depth: 3
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

## Install and Load Packages 
The following code chunk installs the required R packages and loads them onto RStudio environment. sf, an R package specially designed to handle geospatial data in simple feature objects.

```{r load libraries}
packages = c('tidyverse', 'sf', 'tmap', 'lubridate', 'clock',
             'sftime', 'rmarkdown')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p,character.only = T)
}
```

## Importing wkt Data
Well-known text (WKT) is a human readable representation for spatial objects like points, lines, or enclosed areas on a map. 

In the code chunk below, `read_sf()` of *sf* package is used to parse School.csv into R as an sf data.frame.

```{r import data}
schools <- read_sf('data/Schools.csv',
                   options= 'GEOM_POSSIBLE_NAMES=location')
buildings <- read_sf('data/Buildings.csv',
                   options= 'GEOM_POSSIBLE_NAMES=location')
apartments <- read_sf('data/Apartments.csv',
                   options= 'GEOM_POSSIBLE_NAMES=location')
employers <- read_sf('data/Employers.csv',
                   options= 'GEOM_POSSIBLE_NAMES=location')
restaurants <- read_sf('data/Restaurants.csv',
                   options= 'GEOM_POSSIBLE_NAMES=location')
pubs <- read_sf('data/Pubs.csv',
                   options= 'GEOM_POSSIBLE_NAMES=location')
```

## Data Overview
After importing the data file into R, it is important for us to review the data object.

```{r print schools}
print(schools)
```

```{r print buildings}
print(buildings)
```

## Plotting Building Footprint Map: Tmap
The code chunk below plots the building polygon features by using `tm_polygon()`.

```{r tmap}
tmap_mode('view')
tm_shape(buildings) +
  tm_polygons(col = 'grey60',
              size = 1,
              border.col = 'black',
              border.lwd = 1)
tmap_mode('plot')
```

## Plotting A Composite Map

```{r composite, fig.width= 10, fig.height= 12}
tm_shape(buildings)+
tm_polygons(col = "grey60",
           size = 1,
           border.col = "black",
           border.lwd = 1) +
tm_shape(employers) +
  tm_dots(col = '#003366', size = 0.5)
```

*The Task:*
Plot a composite map by combining buildings, apartments, employers, pubs, restaurants, and schools.

```{r composite2, fig.width= 10, fig.height= 12}
tm_shape(buildings) +
  tm_polygons(col = "grey60",
           size = 1,
           border.col = "black",
           border.lwd = 1) +
  tm_shape(employers) +
  tm_dots(col = "#003366", size = 0.5) +
  tm_shape(apartments) +
  tm_dots(col = '#daa520', size = 0.3, alpha= 0.7 ) +
  tm_shape(restaurants) +
  tm_dots(col = '#ff7f50', size = 0.5) +
  tm_shape(pubs) +
  tm_dots(col = '#00ff00', size = 0.5) +
  tm_shape(schools) +
  tm_dots(col = '#20b2aa', size = 0.5)
  
```


```{r import logs, eval=FALSE}
logs <- read_sf('data/ParticipantStatusLogs1.csv',
                   options= 'GEOM_POSSIBLE_NAMES=currentLocation')
```

```{r prepare logs, eval=FALSE}
logs_selected <- logs %>%
  mutate(Timestamp = date_time_parse(timestamp,zone = '',
                                     format = '%Y-%m-%dT%H:%M:%S')) %>%
  mutate(day = get_day(Timestamp)) %>%
  filter(currentMode == 'Transport')
```

```{r save RDS, eval=FALSE}
write_rds(logs_selected, 'data/rds/logs_selected.rds')
```

```{r read rds}
logs_selected <- read_rds('data/rds/logs_selected.rds')
```

```{r haxegons}
hex <- st_make_grid(buildings, 
                    cellsize=100, 
                    square=FALSE) %>%
  st_sf() %>%
  rowid_to_column('hex_id')
plot(hex)
```

```{r}
points_in_hex <- st_join(logs_selected, 
                         hex, 
                         join= st_within) %>%
  st_set_geometry(NULL) %>%
  count(name = 'pointCount', hex_id)
head(points_in_hex)
```

```{r}
hex_combined <- hex %>%
  left_join(points_in_hex, 
            by = 'hex_id') %>%
  replace(is.na(.), 0)
```

```{r, fig.width= 10, fig.height= 12, preview= TRUE}
tm_shape(hex_combined %>%
           filter(pointCount > 0))+
  tm_fill("pointCount",
          n = 8,
          style = "quantile") +
  tm_borders(alpha = 0.1)
```

```{r}
logs_path <- logs_selected %>%
  group_by(participantId, day) %>%
  summarize(m = mean(Timestamp), 
            do_union=FALSE) %>%
  st_cast("LINESTRING")
```

```{r fig.width= 10, fig.height= 12}
logs_path_selected <- logs_path %>%
  filter(participantId == 0)
tm_shape(buildings)+
tm_polygons(col = "grey60",
           size = 1,
           border.col = "black",
           border.lwd = 1) +
  tm_shape(logs_path_selected) +
  tm_lines(col = '#ffa500')
```


