---
title: "Take-home Exercise 6"
categories: Take-home Exercise
description: |
  Putting Visual Analytics into Practical Use: 
  VAST Challenge 2022, Challenge 1: Demographics and Relationships.
author:
  - name: Leslie Long Nu
    url: https://www.linkedin.com/in/leslielongnu/
    affiliation: SMU, MITB
    affiliation_url: https://scis.smu.edu.sg/master-it-business
date: 2022-06-05
preview:
output:
  distill::distill_article:
    code_folding: show
    self_contained: false
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cho = TRUE, eval = TRUE, message = FALSE, warning = FALSE, fig.retina = 3)
```

# 1. Overview
## 1.1 Introduction
With reference to Challenge 1 Question 2 of [VAST Challenge 2022](https://vast-challenge.github.io/2022/), this take-home exercise will reveal  the patterns of community interactions of the city of Engagement, Ohio USA by using social network analysis approach.

*Challenge 1: Demographics and Relationships*

*Consider the social activities in the community. What patterns do you see in the social networks in the town? Describe up to ten significant patterns you observe, with evidence and rationale. *

## 1.2 Approach 
This take-home exercise aims to reveal the social activities in the community by analyzing the social relationship information in the social network file, and demographics of the participants in participants file. 

The scope of this analysis is limited to the social activities happened during weekdays in Mar 2022, at the start of the study. 

# 2. Data Preparation
## 2.1 Install and Load Packages 
The following code chunk installs the required R packages and loads them onto RStudio environment. The loaded packages including four network data modeling and visualization packages, namely [igraph](https://igraph.org/r/), [tidygraph](https://cran.r-project.org/web/packages/tidygraph/index.html), [ggraph](https://cran.r-project.org/web/packages/ggraph/index.html) and [visNetwork](https://datastorm-open.github.io/visNetwork/), as well as some time data wrangling packages such as [tidyverse](https://www.tidyverse.org/) and [lubridate](https://lubridate.tidyverse.org/). 

```{r load libraries, results='hide'}
packages = c('igraph', 'tidygraph', 'ggraph', 'visNetwork',
             'lubridate', 'tidyverse', 'graphlayouts',
             'zoo')
for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```

## 2.2 Import Datasets

**Edges Data**

The edges data, or the link data, contains information on the interaction of the objects (nodes) in the context. In this take-home exercise, the edges data is the social network file that contains information on time and social activities initiated and received by the participants. 
The social network data is imported using the following code chunk: 

```{r import social, eval=FALSE}
social <- read_csv('rawdata/SocialNetwork.csv')
```

**Nodes Data**

The nodes data in this context is the participants file that contains information on the demographics of the participants, such as their household sizes, ages, education level, interest group and etc. 
The social network data is imported using the following code chunk: 

```{r import participants}
participants <- read_csv('data/Participants.csv')
```

## 2.3 Data Wrangling
The following code chunk is used to extract social activities data for the month of Mar 2022.

```{r social Mar, eval=FALSE}
social <- social %>%
  mutate(yearmonth = as.yearmon(timestamp)) %>%
  filter(yearmonth == 'Mar 2022')
```

The following code chunk is used to identify the day of the week of the social interaction. 

```{r day of week, eval=FALSE}
social <- social %>%
  mutate(dow = wday(timestamp))
```

The `wday()` component of a POSIXct object is the numeric weekday, ranging from 0-6 starting on Sunday. Therefore, the following code chunk identifies whether the interaction happened on a weekday or weekend. 

```{r weekday, eval=FALSE}
social <- social %>%
  mutate(daytype = ifelse(dow==0|dow==7, 'Weekend', 'Weekday')) %>%
  select(1:3, 6)
```

The extracted social file is saved and read as RDS format to avoid uploading large data file to GitHub.

```{r save RDS, eval=FALSE}
saveRDS(social, 'data/social.rds')
```

```{r read rds}
social <- readRDS('data/social.rds')
```

The following code chunk cleans the participants file by renaming some columns and values for easy of reading. 

```{r rename participants}
# rename columns
participants <- participants %>%
  rename('Participant_ID' = 'participantId', 
         'Household_Size' = 'householdSize', 
         'Have_Kids' = 'haveKids', 
         'Age' = 'age', 
         'Education_Level' = 'educationLevel', 
         'Interest_Group' = 'interestGroup', 
         'Joviality' = 'joviality')

#rename value 
participants$Education_Level <- sub('HighSchoolOrCollege', 
                                    'High School or College',
                                    participants$Education_Level)
```

Age variable is binned with the following code chunk:

```{r bin age}
# check min and max ages 
summary(participants$Age)

# binning
brks <- c(17, 20, 25, 30, 35, 40, 45, 50, 55, 60)
grps <- c('20 & Below', '21-25', '26-30', '31-35', '36-40', '41-45', 
          '46-50', '51-55', '56-60')

participants$Age_Group <- cut(participants$Age, breaks=brks, labels = grps)
```

The following code chunk bins joviality variable:

```{r bin joviality}
brks <- c(0, 0.25, 0.5, 0.75, 1)
grps <- c('Low', 'Medium to Low', 'Medium to High', 'High')

participants$Joviality_Group <- cut(participants$Joviality, breaks=brks, labels = grps)

```

# 3. Network Visualizations and Insights
The following code chunk checks the composition of social interactions between weekday and weekend. Surprisingly, there are predominately more interaction on weekdays then weekends, weekend social interactions accounts for only 12% of the total, less than half of the average interactions (2/7 = 28.6%). This could be because that as we have found out in the previous take-home exercise that many participants have 2 to 3 recreation activities on weekdays. Another possible reason could be that the social interactions includes interactions in workplaces as well. 

```{r check %}
check <- social %>%
  group_by(daytype) %>%
  summarise(percent = n()/nrow(.))

check
```

## 3.1 Who are the social butterflies at work?
The following code chunk filters the weekday interactions and aggregates the social interactions by to and from participants. due to the nature of high frequency workplace interactions, weight less than 20 are eliminated to reveal the pattern of participants having more than daily interactions. 

```{r aggregate social}
social_edges_weekday <- social %>%
  filter(daytype == 'Weekday') %>%
  group_by(participantIdFrom, participantIdTo) %>%
  summarise(Weight = n()) %>%
  filter(Weight > 20,
         participantIdFrom != participantIdTo) %>%
  ungroup()

```

The following code chunk first filters out irrelevant nodes in partifipants file, then creates a tbl_graph item using `as_tbl_graph()`:

```{r tbl_graph weekday}
# filter irrelevant nodes 
participants_weekday <- participants %>% 
  filter(Participant_ID %in% social_edges_weekday$participantIdFrom |
           Participant_ID %in% social_edges_weekday$participantIdTo)

#tbl_graph
social_graph_weekday <- graph_from_data_frame(social_edges_weekday,
                                      vertices = participants_weekday) %>%
  as_tbl_graph()
social_graph_weekday
```

The output reveals that there are 790 nodes (participants) and 2302 links. 

The following code chunk plots the social network graph using `ggraph()`:

```{r education, fig.width=12, fig.height= 10}
edu_lvl <- c('Low', 'High School or College', 'Bachelors', 'Graduate')

social_graph_weekday %>%
  mutate(Centrality = centrality_betweenness()) %>%
  ggraph(layout = 'stress') +
  geom_edge_link(aes(width = Weight,
                     alpha = 0.6)) +
  scale_edge_alpha(guide = 'none') +
  scale_edge_width(range = c(0.1, 1), 
                   guide = 'none') +
  geom_node_point(aes(color= factor(Education_Level, levels = edu_lvl),
                      alpha = 0.2,
                      size = Centrality)) +
  scale_alpha(guide = 'none') +
  scale_color_discrete(name= 'Education Level') +
  geom_node_text(aes(filter=Centrality > 40000, 
                     label = name),
                 repel = TRUE) +
  labs(title = 'Who are the social butterflies at work?',
       subtitle = 'Weekday Social Network Analysis, Engagement City, Ohio, USA',
       caption = 'Source: Vast Chellenge 2022') +
  theme(legend.position = 'bottom',
        panel.background = element_blank(),
        legend.key = element_rect(fill= NA),
        plot.title = element_text(size = 22),
        plot.subtitle = element_text(size = 16),
        plot.caption = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.text = element_text(size= 12))


```

**Insights**

From the plot, we can derive the following insights:

* There are two types of people in terms of social interactions during weekdays(at work), one that have high interactions with many others on a daily basis (*social butterflies*), and another that have closed-loop high interactions with a small group of people. 
* There are no clear patterns of participants only having high social interactions with people of the same educational background, even for those in small closed-loop social groups. In fact, the plot reveals that participants interact with each other on weekdays/at work regardless of their educational levels. 
* The labelled nodes are the participants that have the highest centralizes on weekdays, namely participants 991, 469, 359, 274, 362, 408, 984, 596, 565 and 544. These participants are closely related to many others and are likely celebrities or opinion leaders of the community. 
* From take-home exercise 1 we learnt that the biggest group of education level is high school or college. However, from the plot we can tell that other than high school and college participants, graduates are highly sociable as well, despite their relatively small group size. 

## 3.2 Who are the most sociable participants?

From the code chunk below, we take a closer look at the most sociable participants at work (on weekdays):

```{r most social}
top10 <- participants_weekday %>%
  filter(Participant_ID == 991 |
           Participant_ID == 469 |
           Participant_ID == 359 |
           Participant_ID == 274 |
           Participant_ID == 362 |
           Participant_ID == 408 |
           Participant_ID == 984 |
           Participant_ID == 596 |
           Participant_ID == 565 |
           Participant_ID == 544)

top10 %>% select(1, 3, 5, 9)
```

From the table above, we can tell that the top 10 sociable participants at work do have some similarities: 

* Most of them do not have kids
* Most of them have high joviality index 

## 3.3 Are sociable participants initiating or receiving interactions?
We take a closer look at the top participants initiating (from) interactions and the top participants receiving (to) interactions. 

The following code chunk locates the top 10 initiator of the weekday social interactions. 

```{r top from }
initiation <- social_edges_weekday %>%
  group_by(participantIdFrom) %>%
  summarise(Initiation = sum(Weight)) %>%
  arrange(desc(Initiation)) %>%
  filter(row_number() <= 10)

top_initiation <- list(initiation$participantIdFrom)
```

The following code chunk locates the top 10 receiver of the weekday social interactions. 

```{r top to}
receiver <- social_edges_weekday %>%
  group_by(participantIdTo) %>%
  summarise(Receiving = sum(Weight)) %>%
  arrange(desc(Receiving)) %>%
  filter(row_number() <= 10)

top_receiver <- list(receiver$participantIdTo)
```

Comparing the list, we realized that the top receivers are also the top initiators of social interactions. 

```{r same}
top_initiation 
top_receiver
```

The social interactions for these participants are extracted and prepared using below code chunk. 

```{r tbl_graph most}
# filter relevant data on the top 10 initiation and receiving participants
social_edges_weekday_most <- social_edges_weekday %>%
  filter(participantIdFrom %in% initiation$participantIdFrom |
           participantIdTo %in% initiation$participantIdFrom)

participants_most <- participants %>%
  filter(Participant_ID %in% social_edges_weekday_most$participantIdFrom |
           Participant_ID %in% social_edges_weekday_most$participantIdTo)
```

The following code chunk prepares the nodes and edges with 'id', 'from', 'to' and 'label' columns and plots the network graph using `vizNetwork()`: 

```{r visnetwork}
participants_most <- participants_most %>%
  rename(group = Joviality_Group,
         id = Participant_ID)

social_most <- social_edges_weekday_most %>%
  rename(from = participantIdFrom,
         to = participantIdTo)

visNetwork(participants_most,
           social_most, 
           main = "Are you in the circles of the most socially active people?",
           submain = 'City of Engagement, Ohio, USA',
           footer = 'Source: VAST Challenge 2022')%>%
  visIgraphLayout(layout = 'layout_with_fr') %>%
  visOptions(highlightNearest = TRUE,
              nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 1234)
```


**Insights**

* It is interesting to note that the 10 most initiator and receivers of social interactions form 7 closed clusters of interactions with others, meaning that some of them are also closely related. As we are looking at weekday social interactions, some of them are likely colleagues, or even family members. 
* It is observed that there are a mixture of joviality groups in terms of the most involved people in their social lives, although they themselves are mostly high joviality levels. 
* By zooming in the graph and selecting particular participant Id, we are able to look at these 10 people's social network closely. 






