---
title: "Take-home Exercise 6"
categories: Take-home Exercise
description: |
  Putting Visual Analytics into Practical Use: 
  VAST Challenge 2022, Challenge 1: Demographics and Relationships.
author:
  - name: Leslie Long Nu
    url: https://www.linkedin.com/in/leslielongnu/
    affiliation: SMU, MITB
    affiliation_url: https://scis.smu.edu.sg/master-it-business
date: 2022-06-05
preview:
output:
  distill::distill_article:
    code_folding: show
    self_contained: false
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

# 1. Overview
## 1.1 Introduction
With reference to Challenge 1 Question 2 of [VAST Challenge 2022](https://vast-challenge.github.io/2022/), this take-home exercise will reveal  the patterns of community interactions of the city of Engagement, Ohio USA by using social network analysis approach.

*Challenge 1: Demographics and Relationships*

*Consider the social activities in the community. What patterns do you see in the social networks in the town? Describe up to ten significant patterns you observe, with evidence and rationale. *

## 1.2 Methodology


# 2. Data Preparation
## 2.1 Install and Load Packages 
The following code chunk installs the required R packages and loads them onto RStudio environment. 

```{r load libraries, results='hide'}
packages = c('igraph', 'tidygraph', 'ggraph', 'visNetwork', 
             'lubridate', 'clock', 'tidyverse', 'graphlayouts')
for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```

## 2.2 Import Datasets






